AGB Stars in the IR: Physical Characteristics, Galactic Spatial Distribution, and Chemical Impact of the Milky Way AGB Population

### 2015-05-05 paper1/paper1_data.tex
ZI edits May 5

\section{Data Sources and Reduction}
\label{sec:data}

\subsection{Data Sources}
\label{sec:sources}
In this study, we rely heavily on data from the \allwise\, extension of the \wise\, survey, combining data from the initial All-Sky Data Release, the 3-band cryogenic data release, and the NEOWISE post-cryogenic data release \citep{2013wise.rept....1C}. The initial \wise\,All-Sky Data Release observed the sky between January and August 2010, observing the sky 1.2 times with four detectors, operating at 3.4, 4.6, 12, and 22$\mu$m. Hereon we refer to \allwise\, photometric bands at [3.4$\mu$m/4.6$\mu$m/12$\mu$m/22$\mu$m] as [$W1/W2/W3/W4$]. The positions of objects in the \wise\, catalog were calibrated to the \twomass\, point source catalog. The 3-band cryogenic data release contains data from $W1$, 2, and 3, and surveyed $30\%$ of the sky between August and October 2010. During the 3-band cryogenic survey, $W1$ and $W2$ operated with nearly the same sensitivity as during the full survey. Warming of the telescope reduced sensitivity in $W3$ and fully saturated $W4$. The NEOWISE post-cryogenic data release contains $W1$ and $W2$ measurements, with sensitivities close to those obtained during the full cryogenic phase. During this phase, \wise\, surveyed $70\%$ of the sky. Data products from the post-cryogenic release included updated instrumental, astrometric, and photometric calibrations and reduction algorithms, resulting in much lower SNR. The overall number of sources compiled into \allwise\, totals over 747.6 million.

>>>>

\section{Input Catalogs and Data Preparation}
\label{sec:data}

Our principal data source is the merged WISE-2MASS catalog. In order to derive
WISE-2MASS color-based selection criteria, we use several training samples
described below: a heterogeneous sample of bright \agb\ stars listed in 
SIMBAD database and more homogneous fainter samples selected using 
optical time-domain surveys OGLE-III and MACHO. The latter samples 
include \agb\ stars from the LMC and SMC and are also used to 
calibrate color-absolute magnitude relations that underlie our distance
estimates. In order to assess what infrared populations could contribute to
contamination of selected \agb\ candidates, we utilize a number of SDSS
catalogs, also described below. We conclude this section by describing
how these catalogs auxiliary catalogs were  positionally merged with the 
WISE-2MASS catalog.

Now rearrange the old text below in these subsections:

\subsection{WISE-2MASS catalog}

\subsection{AGB stars from SIMBAD database}

\subsection{AGB stars from optical time-domain surveys}
A few sentences of introduction...
\subsubsection{MACHO}
\subsubsection{OGLE-III}

\subsection{Imaging and spectroscopic catalogs from SDSS}
A few sentences of introduction...
\subsubsection{Galactic sources} 
\subsubsection{Extragalactic populations}

\subsection{Merged sample}

How was it merged, which further cuts and why...

\subsection{Completeness and contamination of the initial \agb\ sample}

In the next section you are using the \agb\ sample from this section to 
estimate completeness and contamination of color-selected samples. Here
you need to summarize what do you believe completeness and contamination
are for your initial sample constructed using SIMBAD and optical surveys.
In particular, it is important to describe expected distance limits.

\subsection{XXX old stuff from here to the end of this section: Data Sources}
\label{sec:sources}
In this study, we rely heavily on data from the \allwise\, extension of the \wise\, survey, combining data from the initial All-Sky Data Release, the 3-band cryogenic data release, and the NEOWISE post-cryogenic data release \citep{2013wise.rept....1C}. The initial \wise\,All-Sky Data Release observed the sky between January and August 2010, observing the sky 1.2 times with four detectors, operating at 3.4, 4.6, 12, and 22$\mu$m. Hereon we refer to \allwise\, photometric bands at [3.4$\mu$m/4.6$\mu$m/12$\mu$m/22$\mu$m] as [$W1/W2/W3/W4$]. The positions of objects in the \wise\, catalog were calibrated to the \twomass\, point source catalog. The 3-band cryogenic data release contains data from $W1$, 2, and 3, and surveyed $30\%$ of the sky between August and October 2010. During the 3-band cryogenic survey, $W1$ and $W2$ operated with nearly the same sensitivity as during the full survey. Warming of the telescope reduced sensitivity in $W3$ and fully saturated $W4$. The NEOWISE post-cryogenic data release contains $W1$ and $W2$ measurements, with sensitivities close to those obtained during the full cryogenic phase. During this phase, \wise\, surveyed $70\%$ of the sky. Data products from the post-cryogenic release included updated instrumental, astrometric, and photometric calibrations and reduction algorithms, resulting in much lower SNR. The overall number of sources compiled into \allwise\, totals over 747.6 million.

### 2015-05-05 paper1/paper1_intro.tex
ZI edits May 5

- brief intro to Gal structure

- SDSS and other recent results do not go into the plane

- mention a few studies that do (GLIMPSE etc) and ongoing data taking with DECam

- AGB stars: luminous in IR, relatively easy to find; mention MACHO and OGLE studies, 
   refer to Jackson et al. IRAS-based study, lead to WISE

- brief description of WISE

- paper outline

The formation of galaxies like the \mw\, was long thought to be a steady process that created a smooth distribution of stars. This standard view was exemplified by the \cite{1980ApJS...44...73B} and \cite{1989ARA&A..27..555G} models, described in detail by, e.g., \cite{1993ARA&A..31..575M}. It was further motivated by observations of other galaxies, as well as what little information was available from \emph{High Precision Parallax Collecting Satellite} \citep[\emph{HIPPARCOS}, ][]{1984SSRv...39....1K} and smaller pencil-beam surveys. In these, the \mw\, is modeled by three discrete components described by simple analytic expressions: the thin disk, thick disk, and halo.

Review is \citet{IBJ2012} \citep{IBJ2012}

The advent of the \emph{Sloan Digital Sky Survey} \citep[\sdss, ][]{2000AJ....120.1579Y} alleviated these limitations, providing accurate digital multi-band optical photometry across a quarter of the sky, as well as the largest optical spectroscopic catalog thus far known. This new influx of data enabled the development and application of photometric parallax methods, using color-magnitude relations to estimate stellar distances. In turn, this led to the large scale ``tomography" of the \mw\; via stellar distributions in the 7-dimensional space spanned by spatial coordinates \citep{2008ApJ...673..864J}, velocity components \citep{2010ApJ...716....1B}, and metallicity \citep{2008ApJ...684..287I}. The resulting maps revealed rich, complex substructure in the distribution of the \mw's stars \citep[e.g.][]{2000AJ....120..963I,2000ApJ...540..825Y,2001ApJ...554L..33V,2002ApJ...569..245N,2003ApJ...599.1082M,2006ApJ...642L.137B,2006ApJ...651L..29G,2006AJ....132..714V}, deeply shaking the existing view of the Galaxy.

In order to move forward from where \sdss\, tomography left off, we require observations that span an area larger than that of \sdss\, with Galactic objects that can be seen through interstellar dust out to large distances. Stars from the Asymptotic Giant Branch (\agb) are perfect candidates as probes to the \mw. \agb\, stars are represent the last stage of evolution for stars between 0.8 and 8 $M_\odot$ \citep{1983ARA&A..21..271I, 2005ARA&A..43..435H}, so they are bound to reside throughout the galaxy wherever other stars are present. During this phase, they produce substantial stellar winds \citep[$10^{-7} < \dot{M} < 10^{-4} M_\odot$ yr$^{-1}$,][]{2002A&A...391.1053O} rich in SiO and amorphous carbon as they progress through being oxygen-rich to being carbon-rich. These winds collect into circumstellar shells that, when warmed by the stellar photosphere, shine brightly in the near- and mid-infrared (NIR \& MIR respectively).

The \emph{Wide-field Infrared Survey Explorer} \citep[\wise, ][]{2010AJ....140.1868W, 2012wise.rept....1C} is a space-based observatory that has imaged the entire sky in the MIR (3.4, 4.6, 12, and 22$\mu$m). Additionally, \wise\, has been positionally matched to the Two-Micron All-Sky Survey (\twomass), a four-year mission characterizing the full sky in the NIR. Thus, the \wise\, catalog presents with hundreds of millions of sources with photometry of unprecedented sensitivity in the NIR and MIR--ideal for capturing \agb\, stars at Galactic distances. In Section~\ref{sec:data}, we describe in detail the data that we use for our study.

In Section~\ref{sec:criteria}, we describe the color-color criteria used to isolate \agb\, stars in the \wise\, dataset, and the color-magnitude relationships that were derived for these stars from the Large Magellanic Cloud and the \mw\, bulge.
In Section~\ref{sec:distribution}, we describe the spatial density distribution of \agb\, candidates in the Milky Way.
Our conclusions can be found in section~\ref{sec:conclusions}.

>>>>

The structure of the \mw\ holds clues to the processes of formation and evolution of galaxies.
Historical models typically assumed three discrete components described by simple analytic 
expressions: the thin disk, thick disk, and halo \citep{1980ApJS...44...73B, 1989ARA&A..27..555G,1993ARA&A..31..575M}. 
Recent surveys, such as the Sloan Digital Sky Survey \citep[\sdss, ][]{2000AJ....120.1579Y} 
and the Two Micron All Sky Survey \citep[\twomass, ][]{2006AJ....131.1163S}, have provided 
much more detail about these components. For example, SDSS data have constrained stellar distributions 
in the 7-dimensional space spanned by spatial coordinates \citep{2008ApJ...673..864J}, velocity components \citep{2010ApJ...716....1B}, and metallicity \citep{2008ApJ...684..287I}. The resulting maps revealed rich, complex substructure in the distribution of the \mw's stars \citep[e.g.][]{2000AJ....120..963I,2000ApJ...540..825Y,2001ApJ...554L..33V,2002ApJ...569..245N,2003ApJ...599.1082M,2006ApJ...642L.137B,2006ApJ...651L..29G,2006AJ....132..714V}, deeply shaking the older view of a smooth Galaxy.

The large distance limit of about 100 kpc for some halo populations (e.g. RR Lyrae and BHB stars 
with SDSS, XXX Sesar 2010 ref and red giants with 2MASS, XXX Majewski ref), as well as about 10 kpc 
for very numerous main sequence stars, is not reached when studying disk component at low galactic
latitudes. As shown, for example, by Berry et al., the extinction due to interstellar dust limits 
the SDSS sample of main sequence stars to heliocentric distances of only a few kpc. In order
to avoid detrimental effect of dust extinction, stellar samples need to be probed at longer infrared
wavelengths. Several infrared surveys covering the Galactic plane have recently become available
(gives refs for WISE and GLIMPSE, VVV: arXiv:1111.5511).

Among populations suitable for studying Galactic structure using infrared surveys, Asymptotic Giant 
Branch (\agb) stars stand out. \agb\ stars represent the last stage of evolution for stars between 0.8 and 8 $M_\odot$ \citep{1983ARA&A..21..271I, 2005ARA&A..43..435H}. Stars from this mass range can reach the final stages of stellar evolution within the Galactic timescale ($\sim$10 Gyr, \citealt{1983ARA&A..21..271I}) and thus are bound to reside throughout the Galaxy wherever other stars are present. This phase of stellar evolution is marked by two distinct
episodes with different observational characteristics: the early \agb\, phase (E-\agb)  and the thermally-pulsing \agb\, phase (TP). During the thermally-pulsing phase, \agb\, stars produce substantial dust-driven stellar winds \citep[$10^{-7} < \dot{M} < 10^{-4}$ $M_\odot$ yr$^{-1}$,][]{2002A&A...391.1053O} rich in oxides (SiO, Al$_2$O$_3$, etc.) or carbon-rich molecules (SiC, AmC, etc.) The dominant chemical composition is highly dependent upon the metallicity of the host galaxy \citep{2005A&A...434..691M}. High-metallicity galaxies like  the Milky Way have a substantial population of oxygen-rich AGB stars \citep{1985A&A...152L...1H}, whereas low-metallicity galaxies such as the Magellanic clouds are dominated by carbon-rich \agb\ stars \citep{2011AJ....142..103B}. In both cases, the other species of AGB star is rarely seen, as richness in one chemical type (e.g. oxides)  necessitates the almost complete capture of the other chemical type (e.g. carbon) in CO \citep{1983ARA&A..21..271I}.

The dust-rich winds in TP phase  create vast circumstellar shells that are warmed by the stellar photosphere and emit
copiously in the near- and mid-infrared (NIR \& MIR, respectively). Together with high bolometric luminosity
(10$^3$--10$^4$ \Mo), this redistribution of the output radiation to the infrared wavelength makes \agb\ stars
excellent disk probe when infrared survey data are available. Indeed, they were detected all the way to the 
Galactic center even with the IRAS survey \citep{2002MNRAS.337..749J}. Such disk studies can now be significantly
improved thanks to the much more sensitive WISE survey.

The \emph{Wide-field Infrared Survey Explorer} \citep[\wise, ][]{2010AJ....140.1868W, 2012wise.rept....1C}, is a space observatory that has imaged essentially the entire sky in the MIR (3.4, 4.6, 12, and 22 $\mu$m). \wise\, catalog has been positionally matched to the \twomass\ catalog, with the matched catalog listing NIR and MIR 7-band photometry for
hundreds of millions of sources. Given the depths of the two surveys, this catalog should contain AGB stars to many
kpc beyond the Galactic center. Here we develop selection methods for \agb\ stars using WISE and 2MASS data,
and analyse the resulting samples.

In Section~\ref{sec:data}, we describe in detail the WISE, 2MASS and other auxiliary data used in our study. 
In Section~\ref{sec:criteria}, we use samples of known Galactic and Magellanic \agb\, stars to derive 
WISE-2MASS color-based selection criteria, and calibrate color-absolute magnitude relations. 
In Section~\ref{sec:distribution}, we describe the spatial density distribution of selected \agb\, candidates
from the Milky Way. Our conclusions are summarized in Section~\ref{sec:conclusions}.
